\documentclass{article}

\usepackage[submission]{colm2026_conference}
\usepackage{microtype}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{lineno}

\definecolor{darkblue}{rgb}{0, 0, 0.5}
\hypersetup{colorlinks=true, citecolor=darkblue, linkcolor=darkblue, urlcolor=darkblue}

\title{Beyond PII Removal: Risk-Budgeted Generalization for LLM-Ready Sensitive Text}
\author{Anonymous Authors}

\begin{document}

\ifcolmsubmission
\linenumbers
\fi

\maketitle

\begin{abstract}
Sensitive text corpora are valuable for training, retrieval, evaluation, and
post-training adaptation of foundation-model systems, but common
transformation pipelines still focus on direct personally identifying
information (PII) such as names, emails, phone numbers, and addresses. This
span-level view is necessary but incomplete: a record with no explicit name
may still reveal sensitive attributes through quasi-identifiers such as exact
dates, fine-grained locations, rare occupations, schools, medical or legal
facts, family relations, citizenship cues, or combinations of ordinary details.
We introduce Risk-Budgeted Quasi-Identifier Generalization (RB-QIG), a
lightweight and auditable transformation layer that scores quasi-identifiers
by privacy risk and utility importance, applies the least destructive step in
a pre-specified generalization ladder, and stops when a document-level risk
budget is met. On a controlled synthetic benchmark, RB-QIG balanced reduces
risk-weighted leakage to 23.6\% [22.9, 24.5] while preserving 71.7\% [70.8,
72.4] of utility facts; blanket quasi-identifier redaction removes leakage but
preserves only 43.3\% [39.9, 46.8] of utility facts. On a 100-record English
RAT-Bench pilot, direct redaction leaves 78.0\% [73.4, 82.5] LLM-attacker
risk-weighted leakage and a naive LLM sanitizer leaves 46.0\% [40.1, 52.0],
while RB-QIG balanced reduces leakage to 5.7\% [3.2, 8.8]. A blind public
stress test similarly shows large risk reduction relative to direct redaction,
although public semantic utility is tied with blanket redaction. RB-QIG is not
an anonymization guarantee. It is evidence that responsible data
transformation for foundation-model pipelines should measure what capable
models can still infer, not only which explicit PII spans were removed.
\end{abstract}

\section{Introduction}

Sensitive text is central to many data sources that would make
foundation-model systems more useful: clinical notes, legal decisions,
financial records, support logs, internal workplace messages, student advising
records, and user-assistant conversations. Such data can support
retrieval-augmented generation, model evaluation, safety auditing,
preference-data construction, and domain adaptation. The same data also
contains details that people did not intend to make public. Responsible data
enablement therefore often begins with de-identification: remove names,
replace emails, mask phone numbers, and redact addresses.

This direct-identifier pipeline is important, but it is not sufficient for
LLM-ready text. A sentence such as ``a 34-year-old oncology infusion
pharmacist at a small startup in Asheville called about a brother's rare
condition after an April 2025 surgery'' may contain no name or email after
redaction, yet still expose a highly specific profile. Traditional privacy
work formalized this problem through linkage and quasi-identifiers
\cite{sweeney2002kanonymity,machanavajjhala2007ldiversity,li2007tcloseness,narayanan2008netflix}.
Modern language models make the problem more acute because an attacker need
not match exact database columns: LLMs can combine paraphrase recognition,
world knowledge, and contextual inference to recover sensitive attributes
\cite{staab2023beyond}. Recent text-anonymization benchmarks similarly argue
that anonymization should be evaluated by residual re-identification risk, not
only by span recall \cite{krco2026ratbench}.

The practical question is therefore not just ``did the tool remove obvious
PII?'' but ``what can a capable model still infer, and how much useful meaning
was preserved?'' A conservative answer is to redact every quasi-identifier:
occupations become \texttt{[OCCUPATION]}, locations become
\texttt{[LOCATION]}, dates become \texttt{[DATE]}, and diagnoses become
\texttt{[MEDICAL]}. This can reduce leakage, but it also destroys much of the
information that made the corpus useful. The opposite extreme---direct PII
removal only---preserves utility but leaves large residual inference risk.

We study the space between these extremes through Risk-Budgeted
Quasi-Identifier Generalization (RB-QIG). RB-QIG does not train a model and
does not rewrite whole documents. It takes candidate quasi-identifier spans,
assigns each a privacy-risk score and a utility-importance score, and applies
span-level replacements from a small ladder: keep, generalize once,
generalize further, or redact. The algorithm greedily chooses the next edit
that gives the largest risk reduction per unit of utility loss, stopping when
a document-level risk budget is satisfied. The resulting audit log records
which fact was generalized, which replacement was used, and why.

Our contributions are: (i) a quasi-identifier taxonomy for LLM-ready sensitive
text; (ii) a risk-budgeted generalization policy that operationalizes an
inspectable privacy--utility tradeoff without training or fine-tuning; (iii)
an evaluation protocol combining deterministic leakage metrics, paired
bootstrap intervals, cached LLM attackers, and cached LLM utility judges; and
(iv) an analysis showing that residual leakage often comes from relational or
lexical implications rather than missed exact spans. Our strongest supported
claim is narrow but important: direct PII redaction and naive LLM sanitization
leave substantial residual quasi-identifier risk, while RB-QIG sharply reduces
that measured risk and exposes a useful controlled privacy--utility frontier.

\section{Background and Threat Model}

\paragraph{From tabular quasi-identifiers to text.}
Classical privacy work showed that attributes such as ZIP code, gender, and
birth date can link a record to a person even when names are absent
\cite{sweeney2002kanonymity}. $l$-diversity and $t$-closeness refined this
view by showing that indistinguishability alone may not protect sensitive
attributes \cite{machanavajjhala2007ldiversity,li2007tcloseness}.
Differential privacy offers a different guarantee for statistical releases
\cite{dwork2006calibrating}, but it does not directly solve the instance-level
text-transformation problem studied here: preserve a useful text record while
reducing inferable identifiers inside that record.

\paragraph{Text de-identification.}
Applied de-identification systems often target named entities and regulated
identifier categories. Clinical de-identification has a long history of shared
tasks and neural sequence-labeling systems \cite{uzuner2007deid,dernoncourt2017deidentification}.
Practical tools such as Presidio support PII detection, masking, and
replacement \cite{presidio}; specialized privacy filters provide another local
PII-masking building block \cite{openaiPrivacyFilter}. These systems are
useful, but they are typically optimized for identifiable spans rather than
the residual inference risk created by combinations of demographic,
relational, narrative, and domain facts. TAB provides a legal-domain
anonymization benchmark \cite{pilan2022tab}, and RAT-Bench directly emphasizes
LLM-based re-identification risk from direct and indirect identifiers
\cite{krco2026ratbench}.

\paragraph{LLM privacy leakage.}
LLMs introduce two complementary privacy concerns. Training data may be
memorized or extractable under some conditions \cite{carlini2021extracting,carlini2023quantifying}.
Separately, even when a private fact is not explicit, a model may infer it
from surrounding text. Staab et al. show that LLMs can infer personal
attributes from ordinary text in their setting \cite{staab2023beyond}. RB-QIG
focuses on the second concern at the transformation stage: before using
sensitive text for retrieval, evaluation, or post-training, reduce what a
capable model can infer from the transformed record.

\paragraph{Threat model.}
The attacker receives transformed text and, in the LLM-attacker setting, a
list of benchmark attributes to infer. The attacker returns structured guesses
with short evidence. We report exact recovery, coarse recovery, record
compromise, and risk-weighted leakage. We use only synthetic profiles and
public benchmarks; we do not attempt to identify real people. This threat
model does not cover arbitrary auxiliary databases or determined human
investigators. It is a reproducible stress test for whether a transformation
leaves sensitive attributes inferable.

We distinguish three evaluation settings. In the \emph{oracle/controlled}
setting, ground-truth quasi-identifier spans are known from synthetic data. In
the \emph{target-aware public} setting, RAT-Bench target attributes are used to
find value-revealing spans; this isolates transformation quality when the
sensitive attributes of interest are known. In the \emph{blind public}
setting, the extractor does not see target attributes at extraction time and
is scored against benchmark annotations afterward. The blind setting is
closest to deployment and is where our evidence is most caveated.

\section{Risk-Budgeted Quasi-Identifier Generalization}

\subsection{Taxonomy and Generalization Ladders}

RB-QIG uses typed candidate spans. The taxonomy is intentionally broad because
LLM inference often uses facts that are not direct PII. Table~\ref{tab:taxonomy}
summarizes the categories used in the prototype.

\begin{table}[t]
\centering
\small
\caption{Quasi-identifier categories used by RB-QIG. The examples are illustrative;
the implementation stores type, risk, utility, and a replacement ladder for each
candidate span.}
\label{tab:taxonomy}
\begin{tabular}{lll}
\toprule
Category & Example precise value & Safer generalized value \\
\midrule
Date/age & April 17, 2025; 34 & month/year; 30s \\
Location & Asheville, NC & city in the southeastern U.S. \\
Occupation & oncology infusion pharmacist & clinical pharmacy professional \\
Institution & named school or employer & university; healthcare company \\
Medical/legal fact & rare condition or case event & broader condition/event type \\
Family relation & younger brother under guardianship & dependent family member \\
Demographic cue & citizenship or heritage phrase & broader demographic placeholder \\
Narrative cue & unique life event & coarse role-preserving description \\
\bottomrule
\end{tabular}
\end{table}

A generalization ladder is a short ordered list of replacements. For example,
``Boise, Idaho'' may map to ``a city in the U.S. Mountain West,'' then ``a city
in the western United States,'' and finally \texttt{[LOCATION]}. The ladder is
not learned; it is a transparent policy artifact. This matters for sensitive
data owners because they may want to inspect, revise, or prohibit particular
replacement types before releasing transformed text.

\subsection{Risk and Utility Scores}

Each quasi-identifier $q_i$ receives a privacy-risk score $r_i \in [0,5]$ and
a utility-importance score $u_i \in [0,5]$. Risk reflects specificity,
rarity, sensitivity, and linkability. Utility reflects whether the fact is
needed for the benign task: diagnosis category may be important in a clinical
triage note, while an exact date of birth is usually not. The current prototype
uses transparent heuristics and benchmark metadata rather than learned risk
models. This is a limitation, but it also makes the method auditable.

The document risk score is
\[
R(x) = \sum_i r_i + 0.5 \binom{|\{i:r_i \geq 3\}|}{2},
\]
where the second term approximates combination risk from multiple linkable
high-risk attributes. Candidate edits are evaluated by their risk drop divided
by utility loss. RB-QIG greedily applies the edit with the best ratio until
$R(x)$ falls below the budget or no edits remain. All edits are span
replacements, so the system avoids unconstrained paraphrase hallucination.

\subsection{Baselines}

We compare against five baselines. \emph{None} leaves the record unchanged.
\emph{Direct redaction} removes direct identifiers using regexes and
benchmark-provided direct spans. \emph{Naive LLM direct} asks an LLM sanitizer
to remove personally identifying information while preserving meaning.
\emph{Blanket QI redaction} replaces every quasi-identifier with a typed
placeholder. \emph{Presidio-style pattern redaction} is a practical pattern
baseline included as a lower-bound diagnostic rather than a full production
Presidio deployment. RB-QIG is evaluated with strict, balanced, and
utility-oriented budgets in the synthetic setting, with the balanced policy as
the main public method.

\section{Experiments}

\paragraph{Synthetic control.}
We generate 100 synthetic records across medical, legal, financial, workplace,
and education-support domains. Each record has direct identifiers,
quasi-identifiers, ground-truth attributes, a task label, and utility facts.
Because oracle quasi-identifier spans are known, this setting isolates the
transformation tradeoff: given the sensitive spans, does risk-budgeted
generalization preserve more useful information than blanket redaction?

\paragraph{RAT-Bench pilot.}
We evaluate 100 English difficulty-1 RAT-Bench records \cite{krco2026ratbench}.
The target-aware version uses benchmark target attributes to identify
value-revealing spans and variants. This is a controlled transformation
benchmark, not a deployment claim. We also run a blind extractor over the same
100 records and score it against benchmark quasi-identifiers. A deterministic
backstop for common demographic, citizenship, education, employment, marital,
and race/ethnicity cues improves coverage without extra API calls.

\paragraph{TAB legal screen.}
We run a 30-document TAB ECHR legal-domain screen \cite{pilan2022tab}. The TAB
experiment is intentionally bounded and deterministic. It tests whether the
residual-risk story transfers to another domain, but it is not used as a
headline utility win because the small legal-task utility screen is
inconclusive.

\paragraph{LLM attacker and utility judge.}
For RAT-Bench, a cached LLM attacker receives transformed text and requested
attribute names, then returns structured guesses with evidence. A cached LLM
utility judge scores label preservation, fact preservation, and semantic
utility without rewarding preserved direct identifiers. All API calls use
synthetic or public benchmark records, are cached, and cost approximately
\$1.85 fresh-equivalent in the current workspace artifacts.

\paragraph{Metrics.}
Privacy metrics are direct-identifier leakage, exact attribute leakage, coarse
attribute leakage, record compromise, and risk-weighted leakage. Utility
metrics are utility-fact preservation, label preservation, semantic utility,
annotation-derived quasi-identifier specificity, and token change. We report
paired bootstrap 95\% confidence intervals for headline comparisons.

\section{Results and Analysis}

\input{generated/tables.tex}

\paragraph{Controlled privacy--utility frontier.}
Table~\ref{tab:synthetic} shows the cleanest evidence for the method. Direct
redaction preserves utility but leaves all quasi-identifiers. Blanket
quasi-identifier redaction eliminates measured leakage but preserves only
43.3\% of utility facts. RB-QIG balanced preserves 71.7\% of utility facts
while reducing risk-weighted leakage from 100.0\% to 23.6\%. The paired
bootstrap advantage in utility-fact preservation over blanket redaction is
+28.3 points [25.2, 31.6]. This supports the controlled claim that
risk-budgeted generalization can occupy a useful middle point between direct
redaction and blanket redaction.

\paragraph{Semantic utility is more modest.}
The LLM utility judge in Table~\ref{tab:llm-utility} is more forgiving than the
synthetic fact metric because broad placeholders can still preserve some
document meaning. Under this metric, RB-QIG balanced has only a modest semantic
utility edge over blanket redaction: +2.6 points [0.6, 4.8]. On blind public
RAT-Bench, the same judge finds no meaningful edge over blanket redaction. We
therefore scope the large utility claim to controlled synthetic facts and
report public utility as an unresolved limitation.

\paragraph{Direct redaction leaves high residual LLM risk.}
Table~\ref{tab:ratbench-llm} shows that direct redaction is insufficient on the
RAT-Bench pilot. After direct redaction, the LLM attacker still recovers many
target attributes: 78.0\% risk-weighted leakage. The naive LLM sanitizer
improves over direct redaction but still leaves 46.0\% leakage. RB-QIG
balanced reduces leakage to 5.7\%, a paired drop of 72.2 points [67.0, 77.4]
relative to direct redaction and 40.2 points [33.1, 47.1] relative to the LLM
sanitizer. RB-QIG and blanket redaction are statistically tied on this privacy
metric, so the privacy claim is reduction relative to direct and naive
redaction, not superiority over blanket redaction.

\paragraph{Blind extraction is the deployment bottleneck.}
Table~\ref{tab:blind-public} summarizes the blind public stress test. The raw
blind extractor covers 72.8\% of benchmark QI spans; a generic deterministic
backstop raises coverage to 99.6\% without additional API calls. Under the LLM
attacker, blind-backstop RB-QIG balanced reduces risk-weighted leakage from
78.1\% [73.3, 82.9] for direct redaction to 6.4\% [3.9, 9.1], and is
statistically tied with blanket redaction. Public semantic utility is also
tied with blanket redaction, so the blind setting should be read as evidence
for residual-risk reduction and as evidence that robust utility-preserving
quasi-identifier extraction remains open.

\paragraph{Cross-domain and practical-baseline diagnostics.}
The TAB screen provides deterministic cross-domain support: direct redaction
leaves 99.8\% [99.4, 100.0] risk-weighted leakage, while RB-QIG balanced lowers
it to 12.3\% [9.2, 15.7]. Its legal-task utility screen is inconclusive and
therefore not used as a positive utility result. The Presidio-style
pattern-only baseline gives a practical lower-bound diagnostic: on RAT-Bench it
leaves 37.0\% [28.0, 46.0] direct-identifier leakage and 85.0\% [80.6, 89.0]
deterministic risk-weighted leakage; on TAB it misses legal names and case
identifiers, leaving 100.0\% direct-identifier leakage. This makes the
oracle-assisted direct baseline conservative rather than weak.

\paragraph{Failure taxonomy.}
Residual RB-QIG leaks are mostly relational or lexical variants: widowhood from
bereavement cues, sex from gendered language, education from school-name
mentions, citizenship from non-citizen phrasing, and employment or armed-forces
status from occupation context. These failures are not just missed exact spans;
they are cues that a model can combine. This supports the broader argument
that data transformation should evaluate inferability, not only span
categories.

\section{Limitations and Ethics}

RB-QIG is not an anonymization guarantee and should not be used as a legal or
compliance claim. The main public benchmark extraction is target-aware, so it
evaluates transformation behavior under known benchmark attributes rather than
fully blind deployment. Blind public and synthetic diagnostics suggest that
utility-aware conversion and deterministic backstops help, but robust blind
quasi-identifier extraction remains unresolved. LLM attackers and judges are
low-cost automated evaluators and may under- or over-estimate real adversaries
or utility needs. Our risk and utility scores are transparent heuristics, not
universal privacy estimates. All experiments use synthetic or public benchmark
data. We do not attempt to identify real people or infer private attributes
about individuals.

\section{Conclusion}

Direct PII removal is an incomplete transformation for LLM-ready sensitive
text. RB-QIG demonstrates that a small, auditable risk-budgeted generalization
layer can dramatically reduce measured residual inference risk while
preserving useful semantic content better than blanket redaction in controlled
settings. The main open problem is robust blind quasi-identifier extraction and
rewriting that preserves task-relevant utility while handling relational,
lexical, and world-knowledge inference cues. More broadly, responsible data
enablement should evaluate what capable models can still infer, not only what
explicit PII spans were removed.

\appendix
\section{Reproducibility Notes}

The artifact package is generated from local scripts under \path{src/} and
public/synthetic data only. The main synthetic control is
\path{results/synthetic_100/}; the public RAT-Bench pilot is
\path{results/ratbench_d1_api_100/}; and the blind public stress test is
\path{results/ratbench_d1_blind_backstop_v2_budgetfix_api_100/}. Bootstrap
intervals are generated by \path{src/bootstrap_results.py}, and paper tables
by \path{src/make_paper_tables.py}. The qualitative appendix is generated by
\path{src/make_qualitative_appendix.py}; headline claim checks are generated
by \path{src/make_claim_audit.py}. Cached API usage in the workspace is
approximately \$1.85 fresh-equivalent, and all API calls use public or
synthetic benchmark records.

\bibliographystyle{colm2026_conference}
\bibliography{references}

\end{document}
